Distance Sampling in \proglang{R}

### 2016-06-08 paper.Rnw
okay really nearly there...

\section{Introduction}
<<label=setup, echo=FALSE, include=FALSE>>=
options(prompt = 'R> ')
knitr::opts_knit$set(out.format="sweave", cache=TRUE,prompt = TRUE)
render_sweave()
@

@@

We can generate a Q-Q plot and test results using the \code{gof_ds} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
<<label=amakihi-gof, grdevice=NULL>>=
amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
                 key = "hn", adjustment = NULL)
gof_ds(amakihi_hn)
gof_ds(amakihi_hr_obs_mas)
@
We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded. Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models. The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.

\begin{figure}
\begin{center}
<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6>>=
par(mfrow=c(1,2))
gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
@
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

<<label=summary-table, results='asis', echo=FALSE>>=
library("knitr")
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
@

>>>>

\section{Introduction}
<<label=setup, echo=FALSE, include=FALSE>>=
library("knitr")
options(prompt = 'R> ')
knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
knitr::opts_chunk$set(cache=TRUE)
render_sweave()
@

@@

We can generate a Q-Q plot and test results using the \code{gof_ds} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
<<label=amakihi-gof, fig.keep="none">>=
amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
                 key = "hn", adjustment = NULL)
gof_ds(amakihi_hn)
gof_ds(amakihi_hr_obs_mas)
@
We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded. Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models. The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.

\begin{figure}
\begin{center}
<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6, results="hide">>=
par(mfrow=c(1,2))
gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
@
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

<<label=summary-table, results='asis', echo=FALSE>>=
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
@

### 2016-06-08 paper.tex
okay really nearly there...

\documentclass[article]{jss}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{dave@ninepointeightone.net}\\
  URL: \url{http://converged.yt} \\ \\
  Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
  Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
  URL: \url{http://lenthomas.org} \\ \\
  Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
  Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center\\
  National Oceanic and Atmospheric Administration\\
  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \email{Jeff.Laake@noaa.gov}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-detectdists}
\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
\label{fig:detectiondist}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-points-and-lines}
\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
\label{fig:pointslines}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-hn-hr-par-comp}
\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.}
\label{fig:hnhr}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-adjust-mix-comp}
\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
\label{fig:keyadj}
\end{center}
\end{figure}

@@

First we fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
\begin{Schunk}
\begin{Sinput}
R> minke_hn <- ds(minke, truncation = 1.5)
\end{Sinput}
\end{Schunk}
Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvement in AIC.

@@

A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
\begin{Schunk}
\begin{Sinput}
R> minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
\end{Sinput}
\end{Schunk}
Here \code{ds} also fits the hazard-rate model then hazard-rate with a cosine adjustment but the AIC improvement is insufficient to select the adjustment, so the hazard-rate key-only model is returned.

Other adjustment series can be selected using the \code{adjustment=} argument and specific orders of adjustments can be set using \code{order=}. For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
\begin{Schunk}
\begin{Sinput}
R> minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
+                     adjustment = "cos", order = c(1, 2))
\end{Sinput}
\end{Schunk}
Hermite polynomial adjustments use the code \code{"herm"} and simple polynomials \code{"poly"}, adjustment order should be in line with Table \ref{tab:keyadj}.

@@

\code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}. We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}. A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m \citep{Marques:2007ey} can be specified using :
\begin{Schunk}
\begin{Sinput}
R> amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
+                      key = "hr", formula = ~obs)
\end{Sinput}
\end{Schunk}
Note that here, unlike with the minke whale data, \code{ds} warns us that we have only supplied enough information to estimate the detection function (not density or abundance).

While automatic AIC selection is performed on adjustment terms, model selection for covariates must be performed manually. Here we add a second covariate: minutes after sunrise. We will compare these two models further in the following section.
\begin{Schunk}
\begin{Sinput}
R> amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
+                          key = "hr", formula = ~obs + mas)
\end{Sinput}
\end{Schunk}
As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels). However, for point transect studies, probability density function plots give a better sense of model fit than the detection function plots. This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler. The amakihi models included covariates, so the plots show the detection function averaged over levels/values of the covariate. Points on the plot indicate probability of detection for each observation. For the \code{amakihi_hr_obs} model we see fairly clear levels of the observer covariate in the points. Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise (a continuous covariates) to the model.

\begin{figure}
\begin{center}
\includegraphics{paper-minke-amakihi-hn-plot}
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model. For our hazard-rate model of the amakihi data, with observer as a covariate:
\begin{Schunk}
\begin{Sinput}
R> summary(amakihi_hr_obs)
\end{Sinput}
\begin{Soutput}
Summary for distance analysis 
Number of observations :  1243 
Distance range         :  0  -  82.5

@@

We can generate a Q-Q plot and test results using the \code{gof_ds} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
\begin{Schunk}
\begin{Sinput}
R> amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
+                  key = "hn", adjustment = NULL)
R> gof_ds(amakihi_hn)
\end{Sinput}
\begin{Soutput}
Goodness of fit results for ddf object

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.93083  P =  0.003578 
\end{Soutput}
\begin{Sinput}
R> gof_ds(amakihi_hr_obs_mas)
\end{Sinput}
\begin{Soutput}
Goodness of fit results for ddf object

@@

\begin{figure}
\begin{center}
\begin{Schunk}
\begin{Soutput}
Goodness of fit results for ddf object

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.93083  P =  0.003578 
\end{Soutput}
\begin{Soutput}
Goodness of fit results for ddf object

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.15016  P =  0.38908 
\end{Soutput}
\end{Schunk}
\includegraphics{paper-amakihiqq}
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

\caption{Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\label{tab:amakihi}}
\centering
\begin{tabular}[t]{lllrrrr}
\toprule
Model & Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & $\Delta$AIC\\
\midrule
\texttt{amakihi\char`_hr\char`_obs\char`_mas} & Hazard-rate & ~obs + mas & 0.389 & 0.319 & 0.020 & 0.000\\
\texttt{amakihi\char`_hr\char`_obs} & Hazard-rate & ~obs & 0.271 & 0.314 & 0.020 & 1.073\\
\texttt{amakihi\char`_hn} & Half-normal & ~1 & 0.004 & 0.351 & 0.011 & 56.465\\
\bottomrule
\end{tabular}
\end{table}
\section{Estimating abundance and variance}

@@

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by viewing the model summary:
\begin{Schunk}
\begin{Sinput}
R> summary(minke_hn)
\end{Sinput}
\begin{Soutput}
Summary for distance analysis 
Number of observations :  88 
Distance range         :  0  -  1.5

@@

\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Stratum & $\hat{N}$ & $\text{se}(\hat{N}$) & $\text{CV}(\hat{N}$)\\
\midrule
North & 13225.44 & 4966.750 & 0.376\\
South & 3966.46 & 955.962 & 0.241\\
Total & 17191.90 & 5135.586 & 0.299\\
\bottomrule
\end{tabular}
\end{table}
\section{Extensions}

>>>>

\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{Sweave}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{dave@ninepointeightone.net}\\
  URL: \url{http://converged.yt} \\ \\
  Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
  Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
  URL: \url{http://lenthomas.org} \\ \\
  Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
  Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center\\
  National Oceanic and Atmospheric Administration\\
  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \email{Jeff.Laake@noaa.gov}
}
%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/detectdists-1} \end{Schunk}
\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
\label{fig:detectiondist}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/points-and-lines-1} \end{Schunk}
\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
\label{fig:pointslines}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/hn-hr-par-comp-1} \end{Schunk}
\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.}
\label{fig:hnhr}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/adjust-mix-comp-1} \end{Schunk}
\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
\label{fig:keyadj}
\end{center}
\end{figure}

@@

First we fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
\begin{Schunk}
\begin{Sinput}
R> minke_hn <- ds(minke, truncation = 1.5)
\end{Sinput}
\begin{Soutput}
Starting AIC adjustment term selection.
\end{Soutput}
\begin{Soutput}
Fitting half-normal key function
\end{Soutput}
\begin{Soutput}
Key only model: not constraining for monotonicity.
\end{Soutput}
\begin{Soutput}
AIC= 46.872
\end{Soutput}
\begin{Soutput}
Fitting half-normal key function with cosine(2) adjustments
\end{Soutput}
\begin{Soutput}
AIC= 48.872
\end{Soutput}
\begin{Soutput}

Half-normal key function selected.
\end{Soutput}
\end{Schunk}
Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvement in AIC.

@@

A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
\begin{Schunk}
\begin{Sinput}
R> minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
\end{Sinput}
\begin{Soutput}
Starting AIC adjustment term selection.
\end{Soutput}
\begin{Soutput}
Fitting hazard-rate key function
\end{Soutput}
\begin{Soutput}
Key only model: not constraining for monotonicity.
\end{Soutput}
\begin{Soutput}
AIC= 48.637
\end{Soutput}
\begin{Soutput}
Fitting hazard-rate key function with cosine(2) adjustments
\end{Soutput}
\begin{Soutput}
AIC= 50.386
\end{Soutput}
\begin{Soutput}

Hazard-rate key function selected.
\end{Soutput}
\end{Schunk}
Here \code{ds} also fits the hazard-rate model then hazard-rate with a cosine adjustment but the AIC improvement is insufficient to select the adjustment, so the hazard-rate key-only model is returned.

Other adjustment series can be selected using the \code{adjustment=} argument and specific orders of adjustments can be set using \code{order=}. For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
\begin{Schunk}
\begin{Sinput}
R> minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
+                     adjustment = "cos", order = c(1, 2))
\end{Sinput}
\begin{Soutput}
Fitting uniform key function with cosine(1,2) adjustments
\end{Soutput}
\begin{Soutput}
AIC= 48.268
\end{Soutput}
\end{Schunk}
Hermite polynomial adjustments use the code \code{"herm"} and simple polynomials \code{"poly"}, adjustment order should be in line with Table \ref{tab:keyadj}.

@@

\code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}. We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}. A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m \citep{Marques:2007ey} can be specified using :
\begin{Schunk}
\begin{Sinput}
R> amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
+                      key = "hr", formula = ~obs)
\end{Sinput}
\begin{Soutput}
Model contains covariate term(s): no adjustment terms will be included.
\end{Soutput}
\begin{Soutput}
Fitting hazard-rate key function
\end{Soutput}
\begin{Soutput}
AIC= 10778.448
\end{Soutput}
\begin{Soutput}
No survey area information supplied, only estimating detection function.
\end{Soutput}
\end{Schunk}
Note that here, unlike with the minke whale data, \code{ds} warns us that we have only supplied enough information to estimate the detection function (not density or abundance).

While automatic AIC selection is performed on adjustment terms, model selection for covariates must be performed manually. Here we add a second covariate: minutes after sunrise. We will compare these two models further in the following section.
\begin{Schunk}
\begin{Sinput}
R> amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
+                          key = "hr", formula = ~obs + mas)
\end{Sinput}
\begin{Soutput}
Model contains covariate term(s): no adjustment terms will be included.
\end{Soutput}
\begin{Soutput}
Fitting hazard-rate key function
\end{Soutput}
\begin{Soutput}
AIC= 10777.376
\end{Soutput}
\begin{Soutput}
No survey area information supplied, only estimating detection function.
\end{Soutput}
\end{Schunk}
As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels). However, for point transect studies, probability density function plots give a better sense of model fit than the detection function plots. This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler. The amakihi models included covariates, so the plots show the detection function averaged over levels/values of the covariate. Points on the plot indicate probability of detection for each observation. For the \code{amakihi_hr_obs} model we see fairly clear levels of the observer covariate in the points. Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise (a continuous covariates) to the model.

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/minke-amakihi-hn-plot-1} \end{Schunk}
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model. For our hazard-rate model of the amakihi data, with observer as a covariate:
\begin{Schunk}
\begin{Sinput}
R> summary(amakihi_hr_obs)
\end{Sinput}
\begin{Soutput}

Summary for distance analysis 
Number of observations :  1243 
Distance range         :  0  -  82.5

@@

We can generate a Q-Q plot and test results using the \code{gof_ds} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
\begin{Schunk}
\begin{Sinput}
R> amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
+                  key = "hn", adjustment = NULL)
\end{Sinput}
\begin{Soutput}
Fitting half-normal key function
\end{Soutput}
\begin{Soutput}
Key only model: not constraining for monotonicity.
\end{Soutput}
\begin{Soutput}
AIC= 10833.841
\end{Soutput}
\begin{Soutput}
No survey area information supplied, only estimating detection function.
\end{Soutput}
\begin{Sinput}
R> gof_ds(amakihi_hn)
\end{Sinput}
\begin{Soutput}

Goodness of fit results for ddf object

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.93083  P =  0.003578 
\end{Soutput}
\begin{Sinput}
R> gof_ds(amakihi_hr_obs_mas)
\end{Sinput}
\begin{Soutput}

Goodness of fit results for ddf object

@@

\begin{figure}
\begin{center}
\begin{Schunk}

\includegraphics[width=\maxwidth]{figure/amakihiqq-1} \end{Schunk}
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

\caption{Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\label{tab:amakihi}}
\centering
\begin{tabular}[t]{lllrrrr}
\toprule
Model & Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & $\Delta$AIC\\
\midrule
\texttt{amakihi\char`_hr\char`_obs\char`_mas} & Hazard-rate & ~obs + mas & 0.389 & 0.319 & 0.020 & 0.000\\
\texttt{amakihi\char`_hr\char`_obs} & Hazard-rate & ~obs & 0.271 & 0.314 & 0.020 & 1.073\\
\texttt{amakihi\char`_hn} & Half-normal & ~1 & 0.004 & 0.351 & 0.011 & 56.465\\
\bottomrule
\end{tabular}
\end{table}

\section{Estimating abundance and variance}

@@

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by viewing the model summary:
\begin{Schunk}
\begin{Sinput}
R> summary(minke_hn)
\end{Sinput}
\begin{Soutput}

Summary for distance analysis 
Number of observations :  88 
Distance range         :  0  -  1.5

@@

\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Stratum & $\hat{N}$ & $\text{se}(\hat{N}$) & $\text{CV}(\hat{N}$)\\
\midrule
North & 13225.44 & 4966.750 & 0.376\\
South & 3966.46 & 955.962 & 0.241\\
Total & 17191.90 & 5135.586 & 0.299\\
\bottomrule
\end{tabular}
\end{table}

\section{Extensions}
